Pyro-Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-19 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add `\FloatBarrier`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(manuscript): update introductory references
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. @fig-conceptual-model. @fig-model-results. @suppfig-graphical-models.

@@

@@

![Pyro-Velocity accurately predicts cell fate choices and lineage trajectory in human LARRY barcoded hematopoietic cells. Systematic benchmarking of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2). Panels from left to right show on the same SPRING embedding: 1. Cell types; 2. Clonal progression vector field derived based on the observed clonal barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro-Velocity single cell level vector field uncertainty (angular standard deviation from 30 posterior samples); 5. Pyro-Velocity recovered vector field; vectors are colored by the averaged vector field uncertainty; 6. Clonal fate potency score using Cospar24; 7. scVelo recovered latent time; 8. Pyro-Velocity inference of posterior mean shared time per cell. Cosine similarity and Spearman’s correlation for the different methods with the clonal progression and fate potency are reported in the titles.](figures/fig.pdf){#fig-lineage-tracing}

@@

![An example figure (an empty plot)](figures/fig.pdf){#fig-plot}

@@

# Appendix {.appendix}

{{< lipsum 5 >}}

>>>>

RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. The framework has been used to study developmental cell lineage trajectories over a broad spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj], pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors differentiation [@Bergen2020-pj]. Yet, recent state-of-the-art RNA Velocity approaches such as velocyto and scVelo [@La_Manno2018-lj] have several limitations. First, they may predict developmental trajectories that do not exist in nature, as it has been shown in the analysis of fully mature blood cells [@Zheng2017-bz;@Bergen2021-qz]. Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artefactual predictions. In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories (Fig. 1a). Third, most RNA velocity algorithms perform estimates independently for each gene, and therefore cannot exploit the natural interdependence of gene expression. Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time (Fig. 1a). Finally, current RNA velocity frameworks only output point estimates of the velocity vector field and do not provide uncertainty estimation. Importantly, the absence of statistical confidence estimates increases false discoveries as it leaves users without a way to evaluate the robustness of the methods’ outputs. To overcome these limitations, we developed a probabilistic generative model, named Pyro-Velocity, that estimates RNA dynamics and infers cell trajectories using Pyro8, a robust and efficient probabilistic programming language.

@fig-conceptual-model. @fig-model-results. @suppfig-graphical-models.

@@

\FloatBarrier
\newpage

@@

![Pyro-Velocity accurately predicts cell fate choices and lineage trajectory in human LARRY barcoded hematopoietic cells. Systematic benchmarking of cell fate prediction methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated lineages (c, Model 2), or all lineages together (d, Model 2). Panels from left to right show on the same SPRING embedding: 1. Cell types; 2. Clonal progression vector field derived based on the observed clonal barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro-Velocity single cell level vector field uncertainty (angular standard deviation from 30 posterior samples); 5. Pyro-Velocity recovered vector field; vectors are colored by the averaged vector field uncertainty; 6. Clonal fate potency score using Cospar [@Wang2022-xb]; 7. scVelo recovered latent time; 8. Pyro-Velocity inference of posterior mean shared time per cell. Cosine similarity and Spearman’s correlation for the different methods with the clonal progression and fate potency are reported in the titles.](figures/fig.pdf){#fig-lineage-tracing}

\FloatBarrier
\newpage

@@

![Extension of @fig-model-results, Pyro-Velocity (Model 2) outperforms scVelo on PBMC and pancreas datasets. Same format as in @fig-model-results and @suppfig-model-results](figures/fig.pdf){#suppfig-model-2-result}

![Extension of @fig-lineage-tracing, Pyro-velocity (Model 1 for uni-fate data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on scRNA-seq datasets with lineage barcoding information (LARRY dataset). a. SPRING embedding rendering of Cytotrace scores, the title reports the Spearman’s correlation between the Cytotrace and Cospar fate potency scores, b. Pyro-Velocity additional single cell level uncertainty diagnostic plots based on the multi-fate LARRY dataset, from left to right: 1. Standard deviation of posterior samples of shared time per cell, 2. Rayleigh test of posterior samples of vector field per cell (the title reports the false discovery rate after p-value correction for multiple tests across cells), 3. Transcriptional state uncertainty of predicted raw read count per cell. The contour lines show the top 10% of the most uncertain cells (Methods), c. Pairwise Spearman’s correlation between Cytotrace/Cospar fate potency scores and velocity methods for the four different benchmarking datasets selected based on the lineage barcoding information from LARRY data. d. SPRING embedding rendering of posterior mean of shared time across 30 posterior samples, the title shows the Spearman’s correlation of mean shared time with Cytotrace. e. Selection of the top positively correlated genes with the cell shared time based on the posterior mean of spliced expression. f. Phase portraits, rainbow plots, and UMAP rendering with splicing gene expression level for the selected genes in e.](figures/fig.pdf){#suppfig-lineage-tracing}

![Extension of @fig-lineage-tracing presenting an extended comparison of Pyro-Velocity and scVelo on scRNA-seq datasets with lineage barcoding information (LARRY dataset). a. uni-fate Monocyte lineage analysis with Pyro-Velocity Model 2, b. uni-fate Neutrophil lineage analysis with Pyro-Velocity Model 2, c. Neutrophil and Monocyte bi-fate lineage analysis with Pyro-Velocity Model 1, and d. multi-fate lineage analysis with Pyro-Velocity Model 1. The panel descriptions are the same as in @fig-lineage-tracing](figures/fig.pdf){#suppfig-lineage-tracing-2}

![An example figure (an empty plot)](figures/fig.pdf){#suppfig-plot}

@@

\FloatBarrier
\newpage

# Supporting Information {.appendix}
